Case Study: scRNA-seq (Mouse data, scDD method)

### 2018-02-27 datasets/scrnaseq/scrnaseq-mouse-scdd.Rmd
Add summarized benchmark eval; standardize format

For data processing, we follow practices and recommendations from Soneson et al., 2017. Briefly, we limit ourselves to 2-group comparisons. We do not filter genes as filtering may affect the subsequent false discovery control steps.

@@

```{r}
datfile <- paste0(datdir, "mouse_GSE94383.rds")

@@

@@

### scDD: Differences in distributions of gene expression in the 2 conditions

@@

```{r, scdd}
# function to add scDD p-values
compute_scdd_sc <- function(dat){
  # check if already computed
  if (! "scdd_p" %in% colnames(rowData(dat))){
    sce <- SingleCellExperiment(assays=list(count=assays(dat)[["count"]]),
                                colData=data.frame(colData(dat)))
    # scran normalization
    libsize <- scran::computeSumFactors(assay(sce, "count"))
    normcounts(sce) <- GetNormalizedMat(assay(sce, "count"), libsize)

    rowData(dat)$scdd_p <- 
        results(scDD(sce, 
             condition = "group",
             prior_param = list(alpha=0.01, mu0=0, s0=0.01, a0=0.01, b0=0.01), 
             testZeroes = FALSE))$nonzero.pvalue
  }else{
     message("scDD p-values already computed.")
  }
  return(dat)
}

compute_scdd_sc(mousedata)

# save results
saveRDS(mousedata, datfile_m1)
```

@@

# save results
saveRDS(mousedata, datfile_m1)
```

@@

For each covariate, we'll examine the plots for each of the three analysis methods.

@@

```{r, fig.width=10, fig.height=3.2, message=FALSE}
strat_hist(data.frame(rowData(mousedata)), 
           pvalue="scdd_p", covariate="meanExp", maxy=25) 
```

@@

```{r, fig.width=10, fig.height=3.2, message=FALSE}
strat_hist(data.frame(rowData(mousedata)), 
           pvalue="scdd_p", covariate="detection", maxy=25) 
```

### Benchmarking FDR methods

** Still need to clean up from here on; nothing below is eval'd**

First we will adjust for multiple comparisons using the scDD results

```{r, benchmarking, eval=FALSE}

bd <- initializeBenchDesign()
resN <- paste0(resdir, "scrnaseq_wilcox_summarizedBenchmark_",nrow(m_scdd_cov), ".RData")

duration <- NA
if (!file.exists(resN)){
  t1 <- proc.time()
  scrna_sb <- bd %>% buildBench(data=(hscdd1$data%>%select(pval,mean_exp)%>%mutate(effect_size=0, test_statistic=0, SE=0) %>% mutate(ind_covariate=mean_exp)), 
                          ftCols = c("mean_exp"),
                          parallel=TRUE, BPPARAM=multicoreParam)

  save(scrna_sb, file=resN)
  duration <- round((proc.time()-t1)[3]/60,1)
}else{
  load(resN)
}

assayNames(scrna_sb) <- "qvalue"
scrna_sb <- addDefaultMetrics(scrna_sb)

# plot nrejects by method overall and stratified by covariate
rejections_scatter(scrna_sb,supplementary=FALSE)

rejection_scatter_bins(scrna_sb, covariate="ind_covar_N", bins=3,supplementary=FALSE)

# upset plot 
plotFDRMethodsOverlap(scrna_sb, 
                      alpha=0.05, nsets=ncol(scrna_sb),
                      order.by="freq", decreasing=TRUE,
                      supplementary=FALSE)

# properties of method-exclusive discoveries
methods <- c("lfdr", "ihw-a10", "bl-df03", "qvalue", "bh", "bonf")
mcols(scrna_sb)$ind_covariate <- mcols(scrna_sb)$ind_covar_N
plotCovariateBoxplots(scrna_sb, alpha=0.1, nsets=6, methods=methods, trans=log2, maxNum=3)
```

>>>>

For data processing, we follow practices and recommendations from Soneson et al., 2018. Briefly, we limit ourselves to 2-group comparisons. We only filter out genes that are expressed in 
fewer than 1% of cells. 
We will filter out cells with extremely low mapping rate
(below 20%), as these represent poor quality cells that may end up having negative
size factor estimates during normalization. For the same reason, we also filter cells
which have extremely low proportion of genes detected (less than 5%).

@@

```{r}
datfile <- paste0(datdir, "mouse_GSE94383_scdd.rds")

@@

  # exclude cells with extremely low mapping rate or too few expressed features
  # (poor quality cells)

  # exclude cells with extremely low mapping or feature rate (poor quality cells)
  lowMap <- which(metadata(mousedata)$salmon_summary$percent_mapped < 20 |
                  colMeans(assay(experiments(mousedata)[[1]], "count") > 0) < 0.05)
  if (length(lowMap) > 0){
    mousedata <- mousedata[, -lowMap]
    message("Removed ", length(lowMap), " cells with mapping rate below 20%",
            " or fewer than 5% of genes detected.")
  }

@@

@@

```{r, scdd}
# function to add scDD p-values
compute_scdd_sc <- function(dat){
  # check if already computed
  if (! "scdd_p" %in% colnames(rowData(dat))){
    sce <- SingleCellExperiment(assays=list(count=assay(dat, "count")),
                                colData=data.frame(colData(dat)))
    # scran normalization
    libsize <- scran::computeSumFactors(assay(sce, "count"))
    normcounts(sce) <- GetNormalizedMat(assay(sce, "count"), libsize)

    res <- results(scDD(sce, 
             condition = "group",
             prior_param = list(alpha=0.01, mu0=0, s0=0.01, a0=0.01, b0=0.01), 
             testZeroes = TRUE))
    rowData(dat)$scdd_p <- res$nonzero.pvalue
    rowData(dat)$scdd_p_combined <- res$combined.pvalue
  }else{
     message("scDD p-values already computed.")
  }
  return(dat)
}

mousedata <- compute_scdd_sc(mousedata)

# save results
saveRDS(mousedata, datfile)
```

@@

# save results
saveRDS(mousedata, datfile)
```

@@

For each covariate, we'll examine the covariate diagnostic plots.

@@

```{r, fig.width=10, fig.height=3.2, message=FALSE}
strat_hist(data.frame(rowData(mousedata)), 
           pvalue="scdd_p", covariate="meanExp", maxy=16,
           main = "scDD, Covariate 1: Mean Expression") 
```

@@

```{r, fig.width=10, fig.height=3.2, message=FALSE}
strat_hist(data.frame(rowData(mousedata)), 
           pvalue="scdd_p", covariate="detection", maxy=16,
           main = "scDD, Covariate 2: Detection Rate") 
```

## Multiple-Testing Correction

First, we'll create an object of `BenchDesign` class to hold the data and 
add the benchmark methods to the `BenchDesign` object.

```{r, benchdesign, message=FALSE}
bd <- initializeBenchDesign()
```

Now, we're ready to construct the `SummarizedBenchmark` object, which will run
the functions specified in each method (these are actually sourced in from the
                                        helper scripts).

### Covariate one: Mean expression

First we'll include the mean expression covariate.

```{r, sb1, results="hide", message=FALSE}
res1 <- paste0(resdir, "sb_GSE94383_scdd_mean.rds")

if (!file.exists(res1)){
  sb1 <- bd %>% buildBench(data.frame(rowData(mousedata)) %>% 
                                   na.omit() %>%
                                   mutate(pval=scdd_p,
                                          ind_covariate=meanExp), 
                          ftCols = c("meanExp"),
                          parallel=TRUE)

  saveRDS(sb1, file=res1)
}else{
  sb1 <- readRDS(res1)
}
```

Now, we'll repeat the multiple testing correction using the other covariate (detection rate):

```{r, sb2, results="hide", message=FALSE}
res2 <- paste0(resdir, "sb_GSE94383_scdd_det.rds")

if (!file.exists(res2)){
  sb2 <- bd %>% buildBench(data.frame(rowData(mousedata)) %>% 
                                   na.omit() %>%
                                   mutate(pval=scdd_p,
                                          ind_covariate=detection), 
                          ftCols = c("detection"),
                          parallel=TRUE)

  saveRDS(sb2, file=res2)
}else{
  sb2 <- readRDS(res2)
}
```

## Benchmark Metrics

Next, we'll add the default performance metric for q-value assays and 
plot the results. We'll start with covariate one.

First, we have
to rename the assay to 'qvalue'.

```{r, metrics1}
# rename assay to qvalue
assayNames(sb1) <- "qvalue"
sb1 <- addDefaultMetrics(sb1)
```

Now, we'll plot the results.

```{r, plot1, results="hide"}
# plot nrejects by method overall and stratified by covariate
rejections_scatter(sb1,
                   supplementary=FALSE) +
  ggtitle("scDD, Covariate 1: Mean Expression")

rejection_scatter_bins(sb1, covariate="meanExp", bins=4,
                       supplementary=FALSE) +
  ggtitle("scDD, Covariate 1: Mean Expression")

# upset plot 
plotFDRMethodsOverlap(sb1, 
                      alpha=0.05, nsets=ncol(sb1),
                      order.by="freq", decreasing=TRUE,
                      supplementary=FALSE)

# properties of method-exclusive discoveries
methods <- c("ashs", "lfdr", "ihw-a10", "bl-df03", "qvalue", "bh", "bonf",
             "scott-empirical", "scott-theoretical")
mcols(sb1)$ind_covariate <- mcols(sb1)$meanExp
plotCovariateBoxplots(sb1, alpha=0.1, nsets=6, methods=methods, trans=log2, maxNum=3) +
  ggtitle("scDD, Covariate 1: Mean Expression")

```

Next, we'll look at the performance metrics for the other covariate (detection rate).

```{r, metrics2}
# rename assay to qvalue
assayNames(sb2) <- "qvalue"
sb2 <- addDefaultMetrics(sb2)
```

Now, we'll plot the results.

```{r, plot2, results="hide"}
# plot nrejects by method overall and stratified by covariate
rejections_scatter(sb2, supplementary=FALSE) +
  ggtitle("scDD, Covariate 2: Detection Rate")

rejection_scatter_bins(sb2, covariate="detection", bins=4,
                       supplementary=FALSE) +
  ggtitle("scDD, Covariate 2: Detection Rate")

# upset plot 
plotFDRMethodsOverlap(sb2, 
                      alpha=0.05, nsets=ncol(sb2),
                      order.by="freq", decreasing=TRUE,
                      supplementary=FALSE)

# properties of method-exclusive discoveries
methods <- c("ashs", "lfdr", "ihw-a10", "bl-df03", "qvalue", "bh", "bonf",
             "scott-empirical", "scott-theoretical")
mcols(sb2)$ind_covariate <- mcols(sb2)$detection
plotCovariateBoxplots(sb2, alpha=0.1, nsets=6, methods=methods, maxNum=3)+
  ggtitle("scDD, Covariate 2: Detection Rate")

```

# Session Information

```{r}
sessionInfo()
```
